Utility of Diffusion of Innovations for Diagnosing the Spread of Conservation Behaviors

### 2022-01-27 DOI_Review/Manuscript.rmd
text edits

date: "`r format(Sys.time(), '%B %d, %Y')`"
keywords: Community-based conservation, diffusion of innovations, cultural evolution, complex systems
bibliography: [./DOI_Conservation_Review.bib]
abstract: A common goal of conservation practitioners is to design interventions that benefit both people and the environment, and as a result are adopted robustly by local communities. Diffusion of innovations theory (DOI) provides generalized heuristics and models that represent how learned behaviors spread within a population and has been increasingly identified by researchers as useful for understanding conservation adoption. Here, we elaborate on the need for better understanding conservation institutions as a product of individual behavioral adoption and review DOI and its past application to these problems. We follow this with a practical demonstration of how the mathematical models which fit the diffusion process can be used to diagnose and predict the uptake of conservation behaviors even under tightly coupled behavior-environment feedbacks. Data on conservation adoption are generated through an agent-based simulation which highlights our dynamics of interest. Throughout this review, we situate DOI in the broader literature on cultural evolution and use these insights to refine our claims about how diffusion can be productively applied to conservation. Specifically, we show that the social learning mechanism individuals employ can alter the population pattern of adoption and the final proportion of individuals that adopt an intervention. Overall, we provide a quantitative approach for modeling the spread of conservation behaviors and argue for a more complete incorporation of cultural evolutionary theory when considering diffusion of community-based conservation initiatives.

@@

It is becoming increasingly apparent that humanity will not reach the global conservation goals we have set for ourselves through top-down policies alone [@dudley_essential_2018; @mair_achieving_2021; @tittensor_mid-term_2014; @brockington_fortress_2002]. We are thus turning to local communities to self-impose and maintain sustainable limits on resource extraction and even to restore degraded natural areas [@amel_beyond_2017; @catalano_learning_2019]. Community-based conservation projects aim to incentivize these behaviors in rural, and often very poor, communities with little to no external oversight [@western_natural_1994]. While the details of each intervention vary considerably, the general strategy is to closely link community development and conservation goals to promote behavioral changes which are adopted robustly by communities who harvest natural resources to meet their daily needs [@brooks_assessing_2013; @horwich_community_2007]. While some debate still exists, it is largely accepted that empowering communities to self-run conservation institutions and manage environments for multiple interests is a viable method for seeding self-sustaining and ethical protections on natural resources [@turner_global_2007; @berkes_community-based_2007; @adams_biodiversity_2004; @agrawal_enchantment_1999; @yates_purpose_2019].

@@

Decades of research have identified key contextual factors that support the durability of resource safeguarding behaviors in a population, particularly with regard to common-pool resources [@agrawal_common_2001; @baland_halting_1996; @ostrom_governing_1990]. These factors are largely narrative descriptions of the qualities of, and interactions between, resource types, governance structures, and resource users which are most likely to lead to stable and sustainable resource management institutions [@ostrom_general_2009]. Although at least 40 unique factors have been identified as contributing to the success of community-based conservation institutions, well defined social and ecological boundaries, and well connected social systems stand out as major important themes [@ostrom_governing_1990; @ostrom_general_2009; @agrawal_common_2001; @cox_review_2010]. Research on common property additionally recommends that institutions be matched in scale to the underlying ecological processes of the resources they aim to manage [@mckean_common_2000].

Many of the factors that support the long-term viability of conservation institutions are intertwined such that the applicability of one may depend on the level of another and some factors may be inextricably correlated [@agrawal_group_2001; @agrawal_sustainable_2003]. Further, which specific factors are most important is highly specific to a given community or location, suggesting that there is not possibly one unifying panacea for successful conservation institutions [@brooks_how_2012; @geist_proximate_2002; @agrawal_communities_2001]. Thus, even with decades of productive research on sustainable resource management, complete quantitative syntheses and insights have remained elusive.

@@

A major question in implementing these projects is if a specific intervention will indeed be adopted by members of a population and if so, how quickly [@radeloff_hot_2013]? Predicting the final proportion of a population that will adopt an intervention is critical, as many interventions require buy-in from nearly all community members to be effective [@mangubhai_not_2011; @visconti_protected_2019]. By examining sustainable resource management institutions as the emergent outcome of individual behavioral changes, we can leverage insights from other areas of social science to further guide theory on why some institutions emerge and others fail to be adopted long-term [@berkes_rethinking_2004].

@@

# Diffusion of innovations
Promoting the widespread adoption of some novel (to the individual) behavior is not an issue unique to community-based conservation; producers of cigarettes and snack foods, as well as public health professionals, have been grappling with these questions for quite some time. DOI recognizes that this process is primarily governed by the level of uncertainty an individual has about a novel behavior and their personal threshold for the amount of certainty required to make a decision [@rogers_diffusion_2010]. When an individual is introduced to a new behavior, they often have a high degree of uncertainty as to the risks and benefits of adopting and will wait to adopt until that uncertainty is reduced to a tolerable level. The primary way individuals reduce uncertainty about the outcome of adopting a particular behavior is through second-hand observation of early adopters. Therefore, at the core of DOI is the observation that individuals imitate those around them when making decisions about adopting a novel behavior, rather than weighing the pros and cons independently [@henrich_search_2001].

When modifying their beliefs or behaviors, people do not imitate other individuals at random [@bandura_social_1977]. Instead, they preferentially adopt traits by employing a variety of learning biases, such as success-biased learning, where individuals who are perceived as more successful are disproportionately likely to be copied [@boyd_culture_1988; @laland_social_2004]. As a result, the cumulative culture of a population evolves over time as various beliefs and behaviors are unequally copied and their frequency within a population changes. While a full review of cultural evolution is beyond the scope of this paper, the general intuition is that populations carry a variable pool of information which is shaped by events in the lives of individuals such as social observation or natural selection [@mesoudi_cultural_2016]. This information affects individuals’ behaviors, which in turn may affect their environments, and further alter which traits are adopted in the future [@durham_coevolution_1991; @henrich_evolution_2003]. Thus, cultural evolution provides an intuitive way to integrate  individual decision-making, population level patterns, and even population level behavior-environment feedbacks.

@@

When the natural history characteristics of some pathogen are known, the SI, SIR, SIS, & SIRS suite of models reliably give accurate, deterministic predictions about the number of individuals in a population who will become infected and for how long a disease will persist within a population [@huppert_mathematical_2013]. For conservation interventions however, there are no natural history features of a specific intervention, such as the rate or duration of adoption, which can be known ahead of time. Instead, the qualitative insights from DOI and the study of sustainable resource management institutions referenced above aim to increase the rate of spread and reduce the rate of dropout so that the intervention persists in the population and may even spread to new communities. While we can not know exactly what the rates of spread and dropout will be ahead of time, we can use Bayesian inference to probabilistically estimate them once the project has been implemented [@margossian_review_2019; @margossian_differential_2017; @carpenter_predator-prey_2018]. Knowing these parameters may allow us to predict the final rate of uptake from early in the project, identify whether adoption or dropout are limiting the overall rate of uptake, and apply the qualitative insights from DOI and natural resource management institutions in a more targeted fashion.

The uptake of community-based conservation is subject to behavior-environment feedbacks, a fundamental difference from behavioral adoption generally [@liu_complexity_2007]. In the case of marine protected areas for example, success in conserving fish populations may actually disincentivize individuals to continue supporting conservation programs [@christie_marine_2004]. A key question then is whether the diffusion models explained above will still reliably fit the spread of conservation behaviors when those behaviors alter the environment, and changes in the environment alter the payoff structure of various behavioral choices. To answer this question, we use an agent-based simulation to generate data on the spread of a conservation behavior via social learning, contingent on feedbacks with the environment. Then, using just the first 30% of the data, we probabilistically estimate the rate of intervention spread, rate of dropout, and readoption using a Bayesian instantiation of the SIRS model shown in equation 1. We use these estimated parameters to predict the long-term patterns of adoption for the remaining 70% of time steps and compare our predicted values to those observed in the simulation. While the purpose of this example is to test whether the SIRS model can capture our core dynamics of interest, it also serves as an explicit demonstration of how a practitioner or other researchers can directly apply the quantitative process described above. We first briefly describe the simulation below.

>>>>

date: "`r format(Sys.time(), '%B %d, %Y')`"
keywords: Community-based conservation, diffusion of innovations, cultural evolution, complex systems
bibliography: [./DOI_Conservation_Review.bib]
abstract: The study of community-based conservation is plagued by a large number of important variables, nonlinear dynamics, and interactions and feedbacks among predictors. This complexity has made the quantification of causal mechanisms and between-site comparative analyses notoriously difficult. Here, we argue that analyzing the emergence and persistence of community-based conservation institutions at the level of the individual, rather than the entire community, can yield important quantitative insights. We highlight the utility of Diffusion of Innovations theory (DOI) and the broader field of cultural evolution. In particular, the differential models of disease transmission, on which DOI is founded, may provide a productive avenue for quantifying the spread and persistence of community-based conservation institutions. We first simulate data on community adoption of a conservation institution, contingent on feedbacks between individual behavior and environmental processes. We then demonstrate that fitting these data to a probabilistic differential model of diffusion can produce reliable estimates of the rates of adoption, dropout, and long-term uptake from early on in the lifetime of an institution. Specifically, we show that the social learning mechanism individuals employ can alter the population pattern of adoption and the final proportion of individuals that adopt an intervention. Overall, we explore a new quantitative approach for modeling the spread of conservation behaviors using probabilistic differential equations and argue for the incorporation of cultural evolutionary theory into the field.

@@

It is becoming increasingly apparent that humanity will not reach the global conservation targets we have set for ourselves through top-down policies alone [@dudley_essential_2018; @mair_achieving_2021; @tittensor_mid-term_2014; @brockington_fortress_2002]. Conservationists are thus turning to local communities to self-impose and maintain sustainable limits on resource extraction and even to restore degraded natural areas [@amel_beyond_2017; @catalano_learning_2019]. Community-based conservation projects aim to incentivize these behaviors in rural, and often very poor, communities with little to no external oversight [@western_natural_1994]. While the details of each intervention vary considerably, the general strategy is to closely link community development and conservation goals to promote behavioral changes which are adopted robustly by resource dependent communities [@brooks_assessing_2013; @horwich_community_2007]. While some debate still exists, it is largely accepted that empowering communities to self-run conservation institutions is a viable method for seeding self-sustaining and ethical protections on natural resources [@turner_global_2007; @berkes_community-based_2007; @adams_biodiversity_2004; @agrawal_enchantment_1999; @yates_purpose_2019].

@@

Decades of research have identified key contextual factors that support the durability of resource safeguarding behaviors in populations, particularly with regard to common-pool resources [@agrawal_common_2001; @baland_halting_1996; @ostrom_governing_1990]. These factors are largely narrative descriptions of the qualities of, and interactions between, resource types, governance structures, and resource users which are most likely to lead to stable and sustainable resource management institutions [@ostrom_general_2009]. Scholars have identified at least 40 unique factors as contributing to the success of community-based conservation institutions and some overarching patterns have begun to emerge [@ostrom_governing_1990; @ostrom_general_2009; @agrawal_common_2001]. Well-defined social and ecological boundaries, and well-connected social systems particularly stand out as important themes [@cox_review_2010]. Research on common property additionally recommends that institutions be matched in scale to the underlying ecological processes of the resources they aim to manage [@mckean_common_2000].

Many of the factors that support the long-term viability of conservation institutions are intertwined such that the applicability of one may depend on the level of another and some factors may be inextricably correlated [@agrawal_group_2001; @agrawal_sustainable_2003]. Predictors and outcomes of interest commonly exhibit threshold effects, time lags, feedbacks,  and a variety of other nonlinear dynamics, making causal inference through standard regression practices difficult [@folke_socialecological_2007; @levin_social-ecological_2013]. And further, which individual factors are most important is highly specific to a given community or location, suggesting that there is not possibly one unifying panacea for successful conservation institutions [@brooks_how_2012; @geist_proximate_2002; @agrawal_communities_2001]. This complexity of community-based resource management systems, as well as logistical constraints in measuring the sheer number of important variables has made multi-site comparative analyses elusive [@sesmad_social-ecological_2014; @biggs_routledge_2021]. Thus, even with decades of productive research on sustainable resource management, all major syntheses remain descriptive in nature, lacking a robust methodology to quantify causal insights [@frey_synthesis_2017; @mcginnis_social-ecological_2014; @gutierrez_leadership_2011].

@@

A major question in implementing these projects is if a specific intervention will indeed be adopted by members of a population and if so, how quickly [@radeloff_hot_2013]? Predicting the final proportion of a population that will adopt an intervention is critical, as many interventions require buy-in from nearly all community members to be effective [@mangubhai_not_2011; @visconti_protected_2019]. By examining sustainable resource management institutions as the emergent outcome of individual behavioral changes, we can leverage insights from other areas of social science to further guide theory on why some institutions emerge and persist, and others fail to be adopted long-term [@berkes_rethinking_2004].

@@

# Diffusion of innovations
Promoting the widespread adoption of some novel (to the individual) behavior is not an issue unique to community-based conservation; producers of cigarettes and snack foods, as well as public health professionals, have been grappling with these questions for quite some time. DOI recognizes that this process is primarily governed by the level of uncertainty an individual has  about the utility of the novel behavior and their personal threshold for the amount of certainty required to make a decision [@rogers_diffusion_2010]. When an individual is introduced to a new behavior, they often have a high degree of uncertainty as to the risks and benefits of adopting and will wait to adopt until that uncertainty is reduced to a tolerable level. The primary way individuals reduce uncertainty about the outcome of adopting a particular behavior is through second-hand observation of early adopters. Therefore, at the core of DOI is the observation that individuals imitate those around them when making decisions about adopting a novel behavior, rather than weighing the pros and cons independently [@henrich_search_2001].

When modifying their beliefs or behaviors, people do not imitate other individuals at random [@bandura_social_1977]. Instead, they preferentially adopt traits by employing a variety of learning biases, such as success-biased learning, where individuals who are perceived as more successful are disproportionately likely to be copied [@boyd_culture_1988; @laland_social_2004]. As a result, the cumulative culture of a population evolves over time as various beliefs and behaviors are unequally copied and their frequency within a population changes. While a full review of cultural evolution is beyond the scope of this paper, the general intuition is that populations carry a variable pool of information which is shaped by events in the lives of individuals such as social observation or natural selection [@mesoudi_cultural_2016]. This information affects individuals’ behaviors, which in turn may affect their environments, and feedback to further alter which traits are adopted in the future [@durham_coevolution_1991; @henrich_evolution_2003]. Observed variation among groups, such as differences in resource management institutions, are produced by such social and environmental histories, then reinforced through conformist-biased learning where the dominant behavioral trait is disproportionately copied [@henrich_evolution_1998]. Thus, cultural evolution provides an intuitive way to integrate individual decision-making, population level patterns, and even population level behavior-environment feedbacks.

@@

When the natural history characteristics of some pathogen are known, the SI, SIR, SIS, & SIRS suite of models reliably give accurate, deterministic predictions about the number of individuals in a population who will become infected and for how long a disease will persist within a population [@huppert_mathematical_2013]. For a specific conservation interventions however, there is no way to precisely know the natural history features, such as the rate or duration of adoption, ahead of time. Instead, the qualitative insights from DOI and the study of sustainable resource management institutions referenced above aim to increase the rate of spread and reduce the rate of dropout so that the intervention persists in the population and may even spread to new communities. While we can not know exactly what the rates of spread and dropout will be ahead of time, we can use Bayesian inference to probabilistically estimate them once the project has been implemented [@margossian_review_2019; @margossian_differential_2017; @carpenter_predator-prey_2018]. Knowing these parameters may allow us to predict the final rate of uptake from early in the project, identify whether adoption or dropout are limiting the overall rate of uptake, and apply the qualitative insights from DOI and natural resource management institutions in a more targeted fashion.

The uptake of community-based conservation is subject to behavior-environment feedbacks, a fundamental difference from behavioral adoption generally [@liu_complexity_2007]. In the case of marine protected areas for example, success in conserving fish populations may actually disincentivize individuals to continue supporting conservation programs [@christie_marine_2004]. A key question then is whether the diffusion models explained above will still reliably fit the spread of conservation behaviors when those behaviors alter the environment, and changes in the environment alter the payoff structure of various behavioral choices.

To answer this question, we use an agent-based simulation to generate data on the spread of a conservation behavior via social learning, contingent on feedbacks with the environment. Then, using just the first 30% of the data, we probabilistically estimate the rate of intervention spread, rate of dropout, and readoption using a Bayesian instantiation of the SIRS model shown in equation 1. We use these estimated parameters to predict the long-term patterns of adoption for the remaining 70% of time steps and compare our predicted values to those observed in the simulation. While the purpose of this example is to test whether the SIRS model can capture our core dynamics of interest, it also serves as an explicit demonstration of how a practitioner or other researchers can directly apply the quantitative process described above. We first briefly describe the simulation below.
